---
title: "pedana: Create Pedigree Problems, Practice Pedigree Analysis"
author: "Justin Timm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pedana: Create Pedigree Problems, Practice Pedigree Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: "`pedana` is an *R* package providing functions to generate pedigrees and to simulate the transmission of traits of various modes of inheritance. With an included Shiny app, the package offers an environment for creating pedigree problems and exercising pedigree analysis."
---

# pedana

`pedana` is an R package which at its core provides a learning app for 
practicing pedigree analysis. However, to fulfill this function, there are a 
number of interesting functions, i.e. for generating and analysing pedigree 
problems, that can be used outside of the app, too. In this vignette we would 
like to present only the central functions. Further information can be found 
in the documentation of the individual exported functions and in the second 
vignette.

## Pedigree Problems

The basic function of this app is to provide pedigree problems that can be 
clearly assigned to one of the basic modes of inheritance: autosomal dominant, 
autosomal recessive, X-linked dominant, or X-linked recessive.

### Simulate Pedigree Problems

Pedigree problems are simulated using `sim_pedigree_problem()`. The simulation 
time depends on the mode of inheritance in question, additional function 
parameters, and chance. (See the second vignette for a detailed analysis of 
pedana's simulation function regarding simulation time and pedigree structure.) 
You can define the maximum number of iterations and a time limit. However, 
lower values will result in reduced chances of generating a pedigree problem 
successfully. Therefore, you can enforce (and speed up) problem generation by 
setting `force = TRUE` as additional parameter.

```{r sim_pedigree_problem}
library(pedana)

# Simulate a pedigree showing an autosomal recessive trait with
# - max. number of iterations: 20
# - max. iteration duration in seconds: 2
problem_1 <- sim_pedigree_problem(inheritance = "AR",
                                  iter_limit = 10,
                                  time_limit = 2,
                                  seed = 1)
problem_1

# Trying to simulate pedigree showing an X-chromosomal dominant trait
problem_2 <- try(sim_pedigree_problem(inheritance = "XD",
                                      iter_limit = 10,
                                      time_limit = 2, 
                                      seed = 1))

# Try with more time and iterations or use the force
# If force = TRUE the function uses a more purposive method for trait simulation
# and iterates until successful
problem_3 <- sim_pedigree_problem(inheritance = "XD",
                                  force = TRUE,
                                  seed = 1)
problem_3
```

### Plot Pedigree Problems

You can use `plot()` to create the typical pedigree plot. However, this 
function is only a wrapper to the excellent plot method from the *R* package
`kinship2` (for detail, see `plot.ped()` and `kinship2::plot.pedigree()`). 

```{r pedigree-plot, fig.width = 7, fig.height = 4, fig.retina = 2}
plot(problem_1)
```

### Show Solution

To see a solution to an existing pedigree problem, you have two options: First,
you can use `show_solution()` to create an condensed data frame including, 
line-by-line, all hints necessary to determine the present mode of inheritance. 
Second, you can use `verbalise_solution()`, which creates a textual description 
of the solution in German or English. Optional, you can select whether to 
include evidence in the output or not.

```{r solution}
verbalise_solution(problem_1, details = TRUE, l = "en")
show_solution(problem_1, complete = TRUE)
```

### Other Helper Functions

## Practice Pedigree Analysis

To start practicing pedigree analysis run the Shiny app `pedana` via 
`run_pedigree_analysis()`.

```{r run_pedigree_analysis, eval = FALSE}
run_pedigree_analysis()
```

# Acknowledgements

The idea for this package and some of the preliminary work was done during 
my work for the project "Bildungsgerechtigkeit im Fokus" (Educational 
Equity in Focus) at the [University of Duisburg-Essen](https://www.uni-due.de). 
In addition, several people have supported me in various ways in the 
development of this app. First of all, I would like to thank Julia Surmann, who 
did a preliminary work on the development of the pedigree simulation function 
as part of her master's thesis. In addition, I would like to thank Prof. 
Philipp Schmiemannn, Prof. Christian Johannes, Dr. Yvonne Lettmann, Dr. Markus 
Bergmann, Julia Franken, Annika Brechmann, and René Doernemann for great hints 
and ideas from both, genetic and educational perspectives. 

The project "Bildungsgerechtigkeit im Fokus" was financially supported by 
the German Federal Ministry of Education and Research as part of the 
Qualitätspakt Lehre (Quality Pact Teaching) under grant number 01PL16075. 
The responsibility for the content of this publication is held by the author.
