---
title: "Detailed Analysis of pedana's Core Simulation Function"
author: "Justin Timm"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Detailed Analysis of pedana's Core Simulation Function}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: ""
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
ggplot2::theme_set(ggplot2::theme_classic())

if (! requireNamespace("bench", quietly = TRUE) || 
    ! requireNamespace("dplyr", quietly = TRUE) || 
    ! requireNamespace("kableExtra", quietly = TRUE) ||
    ! requireNamespace("ggplot2", quietly = TRUE))
  stop("Can't knit the pedana benchmark vignette, at least one required ",
       "package is missing.")

load("pedana-benchmarks.RData")
```

# pedana

`pedana` is an *R* package providing functions to generate pedigrees for the
typical modes of inheritance: autosomal dominant, autosomal recessive, X-linked
dominant, and X-linked recessive. The core function to use for normal users is
`sim_pedigree_problem()`. Basically, this function is a wrapper to multiple
internal functions generating the pedigree structure and transmitting the trait.
However, pedigree generation does not necessarily complete the pedigree problem
simulation. This depends on whether a conclusively determinable pedigree has
been generated. If not, the simulation must be continued or restarted. For
X-linked modes of inheritance, this is often the case and problem simulation
can take a relatively long time (in some cases longer than one minute).
To address this problem, I have developed specialized simulation functions
for both X-linked modes of inheritance to speed up the problem generation.
This mode can be activated by setting the `force` argument in
`sim_pedigree_problem()` to `TRUE`. An example of this can also be found in the
main vignette.

## Benchmarking and Analysing `sim_pedigree_problem()`

Using package `bench` allows to illustrate the effect of `force` and
`inheritance` by estimating `sim_pedigree_problem()` execution times as a
function of `inheritance` and `force`. In addition, it seems worthwhile to
take a closer look at the structure of the simulated pedigrees to ensure that 
there are no interfering correlations between superficial pedigree features 
(such as trait frequency) and the mode of inheritance.

## Execution Times

```{r, echo = F}
execution_times <- bench_data |> 
  dplyr::filter(force == TRUE) |> 
  dplyr::summarise(mean = mean(time), 
                   sd = sd(time)) |>
  round(digits = 2)

n_data <- NROW(bench_data[bench_data$inheritance == "Autosomal Dominant" &
                            bench_data$force == TRUE, ])
```

In general, the simulation functions are able to generate pedigree problems 
for the learning app at reasonable execution times.^[Simulations were performed 
on a Macbook Air (Retina, 2018). A set of `r n_data` pedigree problems was 
generated for each parameter combination, giving a total of `r n_data * 8` 
pedigrees.] It can be assumed that a mean simulation duration of 
*M* = `r sprintf("%.2f", execution_times$mean)` s 
(*SD* = `r sprintf("%.2f", execution_times$sd)` s) is not perceived to 
distracting by learners. Note: Values of mean and standard deviation are 
estimated based on execution times generated with argument `force` set to 
`TRUE`, as in the app. This must be taken into account when interpreting the 
following table displaying execution times.

```{r bench-table, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
bench_data |>
  dplyr::select(inheritance, force, time) |>
  dplyr::group_by(inheritance, force) |>
  dplyr::summarise(`min` = bench::as_bench_time(min(time)),
                   `mean` = bench::as_bench_time(mean(time)),
                   `75%` = bench::as_bench_time(quantile(time, prob = .75)),
                   `95%` = bench::as_bench_time(quantile(time, prob = .95)),
                   `max` = bench::as_bench_time(max(time)),
                   .groups = "drop_last") |>
  kableExtra::kbl() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

The following results impressively show the advantage of the `force` on the
simulation duration for X-linked modes of inheritance. Since no optimized
functions exist for the autosomal modes of inheritances, the simulation times
do not differ significantly here.

```{r bench-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4, fig.retina = 2}
bench_data |>
  ggplot2::ggplot(ggplot2::aes(x = inheritance, y = time, color = force)) +
  bench::scale_y_bench_time() +
  ggplot2::geom_boxplot() +
  ggplot2::labs(x = "Force Used for Pedigree Simulation",
                y = "Time")
```

## Proportion Affected

The proportion affected (PR) describes the proportion of affected individuals
among all family members:

$$PA = \frac{n_{\text{affected}}}{n_{\text{total}}} $$

The range of possible values can theoretically range from 0 to 1. Nevertheless,
these limits cannot be found among the simulated pedigrees, because, obviously,
they cannot be assigned to one mode of inheritance without doubt.

```{r prop-affected-table, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
bench_data |>
  dplyr::group_by(inheritance) |>
  dplyr::summarise(`min` = min(prop_affected),
                   `25%` = quantile(prop_affected, prob = .25),
                   `mean` = mean(prop_affected),
                   `75%` = quantile(prop_affected, prob = .75),
                   `max` = max(prop_affected),
                   `SD` = sd(prop_affected)) |>
  kableExtra::kbl(digits = 2) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

The results of the simulation indicate that the autosomal modes of inheritance
cannot be differentiated by the proportion of affected individuals, as desired.
Actually, deviations in the proportion are only noticeable in the case of
X-linked recessive inheritance, where pedigrees with few affected individuals
predominate. However, it has not yet been possible to optimise the corresponding
simulation function without accepting other disadvantages.

```{r prop-affected-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4, fig.retina = 2}
ggplot2::ggplot(data = bench_data, ggplot2::aes(x = prop_affected, color = force)) +
  ggplot2::geom_density(alpha = 0.2) +
  ggplot2::facet_wrap(. ~ inheritance, scales = "free_y") +
  ggplot2::scale_x_continuous(limits = c(0, 1)) +
  ggplot2::labs(x = "Proportion Affected (PA)",
                y = "Density")
```

## Gender Ratio

The gender ratio (GR) describes the gender ratio among affected and can range
from -1 to 1:

$$GR = \frac{n_{\text{affected males}} - n_{\text{affected females}}}{n_{\text{affected males}} + n_{\text{affected females}}} $$

The smallest possible value of -1 emerges when
only females are affected. The maximum value of 1 occurs when only males are
affected. A value of 0 represents a pedigree where men and women are affected
equally often.

Analysis of the simulated pedigrees shows that, on average, males and females
are equally affected in the autosomal modes of inheritance. Nevertheless, there
is a considerable variance. This means that the `sim_pedigree_problem()`
function can generate pedigrees with a gender imbalance among affected persons
which can nevertheless be uniquely determined as autosomal. Thus, an imbalance
does not necessarily indicate X-linked inheritance.

```{r gender-ratio-table, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.show="hold", out.width="50%"}
bench_data |>
  dplyr::group_by(inheritance) |>
  dplyr::summarise(`min` = min(gender_ratio),
                   `25%` = quantile(gender_ratio, prob = .25),
                   `mean` = mean(gender_ratio),
                   `75%` = quantile(gender_ratio, prob = .75),
                   `max` = max(gender_ratio),
                   `SD` = sd(gender_ratio),
                   .groups = "drop_last") |>
  kableExtra::kbl(digits = 2) |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

As expected, females are more frequently affected in pedigrees showing an
X-linked dominant inheritance and males are more frequently affected in
pedigrees representing an X-linked recessive inheritance.

```{r gender-ratio-plot, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 4, fig.retina = 2}
ggplot2::ggplot(data = bench_data, ggplot2::aes(x = gender_ratio, color = force)) +
  ggplot2::geom_density(alpha = 0.2) +
  ggplot2::facet_wrap(. ~ inheritance, scales = "free_y") +
  ggplot2::scale_x_continuous(limits = c(-1, 1)) +
  ggplot2::labs(x = "Gender Ratio Among Affected (GR)",
                y = "Density")
```
